---
title: "Fitness Prediction"
author: "Sai"
date: "October 25, 2015"
output: html_document
---

##Background

Using devices such as Jawbone Up, Nike FuelBand, and Fitbit it is now possible to collect a large amount of data about personal activity relatively inexpensively. These type of devices are part of the quantified self movement  a group of enthusiasts who take measurements about themselves regularly to improve their health, to find patterns in their behavior, or because they are tech geeks. One thing that people regularly do is quantify how much of a particular activity they do, but they rarely quantify how well they do it. In this project, your goal will be to use data from accelerometers on the belt, forearm, arm, and dumbell of 6 participants. They were asked to perform barbell lifts correctly and incorrectly in 5 different ways. More information is available from the website here: http://groupware.les.inf.puc-rio.br/har (see the section on the Weight Lifting Exercise Dataset).

##Data

The data for this project come from this source: [http://groupware.les.inf.puc-rio.br/har](http://groupware.les.inf.puc-rio.br/har). The information has been generously provided for use use in this cousera course by the authors, Velloso, E.; Bulling, A.; Gellersen, H.; Ugulino, W.; Fuks, H. They have allowed the use of their paper “Qualitative Activity Recognition of Weight Lifting Exercises. Proceedings of 4th International Conference in Cooperation with SIGCHI (Augmented Human ’13) . Stuttgart, Germany: ACM SIGCHI, 2013.

The training data for this project are available here:

[https://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv](https://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv)

The test data are available here:

[https://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv](https://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv)
```{r}
downloadcsv <- function(url, nastrings) {
    temp <- tempfile()
    download.file(url, temp, method = "curl")
    data <- read.csv(temp, na.strings = nastrings)
    unlink(temp)
    return(data)
}

trainurl <- "https://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv"
train <- downloadcsv(trainurl, c("", "NA", "#DIV/0!"))

testurl <- "https://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv"
test <- downloadcsv(testurl, c("", "NA", "#DIV/0!"))
```
The training data has 19622 observations and 160 features, and the distribution of the five measured stances A,B,C,D,E is:

```{r}
dim(train)
table(train$classe)
```

##Preprocessing

###Partitioning the training set

We separate our training data into a training set and a validation set so that we can validate our model.
```{r}
library(caret)
set.seed(123456)
trainset <- createDataPartition(train$classe, p = 0.8, list = FALSE)
Training <- train[trainset, ]
Validation <- train[-trainset, ]
```

###Feature selection

First we clean up near zero variance features, columns with missing values and descriptive fields.
```{r}
# exclude near zero variance features
nzvcol <- nearZeroVar(Training)
Training <- Training[, -nzvcol]

# exclude columns with 40% or more missing values 
# exclude descriptive columns like name etc
cntlength <- sapply(Training, function(x) {
    sum(!(is.na(x) | x == ""))
})
nullcol <- names(cntlength[cntlength < 0.6 * length(Training$classe)])
descriptcol <- c("X", "user_name", "raw_timestamp_part_1", "raw_timestamp_part_2", 
    "cvtd_timestamp", "new_window", "num_window")
excludecols <- c(descriptcol, nullcol)
Training <- Training[, !names(Training) %in% excludecols]
```

###Model Train

We will use random forest as our model as implemented in the randomForest package by Breiman's random forest algorithm (based on Breiman and Cutler's original Fortran code) for classification and regression.
```{r}
library(randomForest)
rfModel <- randomForest(classe ~ ., data = Training, importance = TRUE, ntrees = 10)
```

###Model Validation

Let us now test our model performance on the training set itself and the cross validation set.

**Training set accuracy**
```{r}
ptraining <- predict(rfModel, Training)
print(confusionMatrix(ptraining, Training$classe))
```
Obviously our model performs excellent against the training set, but we need to cross validate the performance against the validation set and see if we have avoided overfitting.

###Validation set accuracy (Out-of-Sample)

Let us now see how our model performs on the cross validation set that we created from original training set.
```{r}
pvalidation <- predict(rfModel, Validation)
print(confusionMatrix(pvalidation, Validation$classe))
```

###Decision
The cross validation accuracy is 99.5% and the out-of-sample error is therefore 0.5% so our model performs rather good. The expected out-of-sample error is calculated as 1 - accuracy for predictions made against the cross-validation set. Our Test data set comprises 20 cases. With an accuracy above 99% on our cross-validation data, we can expect that very few, or none, of the test samples will be missclassified.

###Test set prediction

The prediction of our algorithm for the test set is:
```{r}
ptest <- predict(rfModel, test)
ptest
```

**Submission:**
```{r}
answers <- as.vector(ptest)

pml_write_files = function(x) {
    n = length(x)
    for (i in 1:n) {
        filename = paste0("problem_id_", i, ".txt")
        write.table(x[i], file = filename, quote = FALSE, row.names = FALSE, 
            col.names = FALSE)
    }
}


pml_write_files(answers)
```
